Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-10 arxiv-version/mistake-lower-bound-for-ignorant-algorithms.tex
add ignorant algorithm analysis

\section{Mistake lower bound for ignorant algorithms}

\begin{lemma}
For any $\gamma \in (0,\frac 1 {160})$,
there exists a set of $N$ examples $x_1, \ldots, x_N$ and vectors $w_1, \ldots, w_N$ ($N = (\Omega(\frac{1}{\gamma}))^{(d-2)/2}$), such that:
$\| x_i \|_2 \leq 1$, $\| w_i \|_2 \leq 1$ for all $i$ in $[N]$, and for all $i, j$ in $[N]$,
\[ w_i \cdot x_j \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases} \]
\label{lem:embed_d_gamma}
\end{lemma}
\begin{proof}
By Lemma 6 of~\cite{Long-1995}, there exists a set of $N = (\frac{1}{2\sqrt{40\gamma}})^{d-2}$ $(d-1)$-dimensional vectors $z_1, \ldots, z_N$ on the unit sphere $\mathbb{S}^{d-2}$, such that
$\theta(z_i, z_j) \geq \sqrt{40 \gamma}$. As $\cos\theta \leq 1-\theta^2/5$ for every $\theta \in [0,\pi]$,
this implies that
\[ z_i \cdot z_j \begin{cases} = 1 ,& i = j, \\ \leq 1- 8\gamma ,& i \neq j. \end{cases} \]

Define $x_i = (\frac12 z_i, \frac12)$, and $w_i = (\frac 12 z_i, -\frac 12(1-4\gamma))$ for all $i$ in $[N]$. It can be easily checked that for all $i$, $\|x_i\| \leq 1$ and $\|w_i\| \leq 1$. In addition,
\[ w_i \cdot x_j = \frac 1 4 z_i \cdot z_j - \frac {1-4\gamma} 4 \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases}\]
\end{proof}

>>>>

\section{Mistake lower bound for ignorant algorithms}
In this section, we prove a mistake lower bound for a family of algorithms called \textit{ignorant algorithms}. 
For simplicity, we assume $(x_t, y_t)\in\mathbb{R}^{d}\times [K]$ in this section. 
As the name suggests, ignorant algorithms ignore the examples on which they make mistakes. 
This assumption seems strong, but as we will explain below, it is actually natural, 
and several recently proposed bandit classification algorithms that achieve the $\sqrt{T}$-regret bound belong to this family, e.g., SOBA~\citep{Beygelzimer-Orabona-Zhang-2017}, OBAMA~\citep{Foster-Kale-Luo-Mohri-Sridharan-2018}. 
We will see that under the $\gamma$-weak linearly separability assumption, 
if a bandit classification algorithm wants to achieve a mistake bound better than $\Theta\left(\min\{\sqrt{T}, (1/\gamma)^{\Theta(d)}\}\right)$, where $d$ is the feature dimension, it must perform some meaningful update when it makes a mistake.

The ignorant algorithm is formally defined as follows.  
\begin{definition} \label{definition:ignorant-algorithm}
    A bandit classification algorithm $\mathcal{A}$ is ignorant if for all $t\in[T]$, 
    $\widehat y_t\in \{1,2\}$ is drawn from some distribution that is jointly determined by 
    \begin{enumerate}
        \item $x_t$
        \item $H_t\triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$, 
        where $1 \leq \tau_1<\ldots <\tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$ 
        are the rounds in which the algorithm predicts correctly (that is, 
        $\{\tau_1, \ldots, \tau_{n_t}\} = \{1\leq \tau<t: \widehat y_\tau=y_\tau\}$). 
    \end{enumerate}
    We use $p_t(\cdot|x_t)\in \{p\in \mathbb{R}_+^2: p^\top\mathbf{1}=1\}$ to denote the distribution $\widehat y_t$ is drawn from. Note that we let the current feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on $x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is only determined by $H_t$. 
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense. Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat y_t \neq y_t$). 
Then at time $t+1$, if it is given the same feature vector $x_{t+1}=x_t$, it will predict the label with the same distribution $p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. 
Below we describe the difficulty of learning from mistakes using a class of natural approaches under the weak separability assumption. 
Consider an approach similar to OBAMA: in each round $t$, 
we create a \textit{convex} loss function $\widehat \ell_t: \mathbb{R}^{K\times d}\rightarrow \mathbb{R}$ that gives a loss value $\widehat \ell_t(W)$ to 
all linear models $W\in\mathbb{R}^{K\times d}$ ($W=[w_1 \ldots w_K]^\top$ for $w_1, \ldots, w_K$ defined in Definition~\ref{definition:linear-separability}). 
The requirement for $\widehat{\ell}_t$ to be convex is due to computational efficiency. 
To let $\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models $\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \geq \ip{w_i}{x_t}, \forall i\in[K]\}$, 
while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller. 
The following fact can be verified when $K\geq 3$, which suggests there is no way to define the loss $\widehat \ell_t$ that satisfies the above properties constraints: 
there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin \mathcal{W}$. 
On one hand, by convexity, we should have $\widehat{\ell}_t(W_2)\leq \max\{\widehat{\ell}_t(W_1), \widehat{\ell}_t(W_3)\}$; 
on the other hand, to penalize $\mathcal{W}$, we should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. 
Clearly, these two requirements contradict each other.

The following theorem states the mistake lower bound for ignorant algorithms. 
\begin{theorem}[Mistake Lower Bound for Ignorant Algorithms]
\label{theorem:ignorant_lower_bound}
    Suppose $\mathcal{A}$ is an ignorant bandit classification algorithm. 
    There exists an adversary that can sequentially choose $(x_1, y_1), (x_2, y_2), \ldots, (x_T, y_T) \in \mathbb{R}^d\times \{1,2\}$, 
    such that $\forall t\in[T], \|x_t\|\leq 1$, $\{(x_t, y_t)\}_{t=1}^T$ is $\gamma$-strongly linearly separable, 
    and the expected number of mistakes made by $\mathcal{A}$ is lower bounded as 
    \begin{align*}
        \Exp\left[ \sum_{t=1}^T \indicator{\widehat y_t \neq y_t} \right] \geq \frac{1}{10}\min\left\{\sqrt{T}, 
        \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\}. 
    \end{align*}
\end{theorem}
Before proving the theorem, we need the following lemma. 
\begin{lemma}
\label{lem:embed_d_gamma}
Let $\gamma \in (0,\frac 1 {160})$. 
There exists vectors $v_1, \ldots, v_N, u_1, \ldots, u_N\in \mathbb{R}^d$, where $N = (\Omega(\frac{1}{\gamma}))^{(d-2)/2}$, such that:
$\| v_i \|_2 \leq 1$, $\| u_i \|_2 \leq 1$ for all $i$ in $[N]$, and for all $i, j$ in $[N]$,
\[ u_i \cdot v_j \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases} \]
\end{lemma}
\begin{proof}
By Lemma 6 of~\cite{Long-1995}, there exists a set of $N = (\frac{1}{2\sqrt{40\gamma}})^{d-2}$ $(d-1)$-dimensional vectors $z_1, \ldots, z_N$ on the unit sphere $\mathbb{S}^{d-2}$, such that
$\theta(z_i, z_j) \geq \sqrt{40 \gamma}$. As $\cos\theta \leq 1-\theta^2/5$ for every $\theta \in [0,\pi]$,
this implies that
\[ z_i \cdot z_j \begin{cases} = 1 ,& i = j, \\ \leq 1- 8\gamma ,& i \neq j. \end{cases} \]

Define $v_i = (\frac12 z_i, \frac12)$, and $u_i = (\frac 12 z_i, -\frac 12(1-4\gamma))$ for all $i$ in $[N]$. It can be easily checked that for all $i$, $\|v_i\| \leq 1$ and $\|u_i\| \leq 1$. In addition,
\[ u_i \cdot v_j = \frac 1 4 z_i \cdot z_j - \frac {1-4\gamma} 4 \begin{cases} \geq \gamma ,& i = j, \\ \leq -\gamma ,& i \neq j. \end{cases}\]
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (PartI: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}. 
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lem:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lem:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\geq 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$. 
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$. 
\ENDFOR

}
\end{algorithmic}
\end{algorithm}
For this strategy, we prove the mistake lower bound of any ignorant algorithm. 
Define the indicators 
\begin{align*}
A_t &= \indicator{\forall \tau\leq t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\leq t, p_\tau(1|x_\tau)\geq 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}. 
\end{align*}
Then we have 
\begin{align}
    \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] 
    &\geq \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
    & \geq \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}A_t\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
    & = \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right] \nonumber \\
    & \geq \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}} B_{t+1}\right], 
    \label{eqn:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use $A_t+B_t\leq 1$; 
in the equality, we use  $B_{t+1}=B_t\indicator{\widehat y_t\neq y_t}$ by $B_t$'s definition; 
in the last inequality we use the fact that when $t\leq N$ and $A_t=1$, it must be $y_t=1$. 
Now, denote $T_1 = \argmin_{\tau} \{ p_\tau(1|x_\tau) \geq 1-q_0 \}$ (if such $\tau$ does not exist or is larger than $\min\{T,N\}$, we simply let $T_1=\min\{T,N\}+1$). 
Then $A_t=1$ for all $t\leq T_1-1$.

Note that $B_{T_1}=1$, and for $t\geq T_1$ we have $\Exp[B_{t+1}=0|B_t=1]=\Pr[\widehat y_t=y_t|B_t=1]=\Pr[\widehat y_t=2|B_t=1]\leq q_0$ because the algorithm is ignorant. Thus the second term in \eqref{eqn:mistake_lower_bound_temp} can be calculated as

\begin{align*}
    \mathbf{E}\left[\sum_{t=1}^{\min\{T,N\}}B_{t+1}\right]\geq \mathbf{E}\left[\sum_{t=T_1}^{\min\{ T,N \}} (1-q_0)^{t-T_1+1} \right]
    = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{\min\{T,N\}-T_1+1}\right]. 
\end{align*}
Combining \eqref{eqn:mistake_lower_bound_temp} and the above inequalities, we get
\begin{align}
    \mathbf{E}\left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right]\geq \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{\min\{T,N\}-T_1+1}\right)\right]. \label{eqn:mistake_lower_bound_temp2}
\end{align}
In the case $T_1\geq \frac{1}{2}\min\{T,N\}+1$, the right-hand side of \eqref{eqn:mistake_lower_bound_temp2} is lower bounded by $\frac{1}{2}q_0\min\{T,N\}=\frac{1}{2}\min\{\sqrt{T}, \sqrt{N}\}$; in the case $T_1< \frac{1}{2}\min\{T,N\}+1$, it is lower bounded by 
\begin{align*}
    \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}\min\{T,N\}}\right)= \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right)\geq \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right)\geq \frac{1}{10}\min\{\sqrt{T}, \sqrt{N}\}.  
\end{align*}
Combining both cases concludes the proof. 
\end{proof}

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning the adversary always set the label to $1$; 
once it sets $y_t=2$ at some $t$, it lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\geq t$. 
Consider $w_2=u_t$ and $w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lem:embed_d_gamma}. 
Then by Lemma~\ref{lem:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \geq \gamma, \forall s \geq t$ 
and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \leq - \gamma, \forall s< t$; 
also $\ip{w_1}{x_s} \leq -\gamma, \forall s \geq t$ 
and $\ip{w_1}{x_s} \geq  \gamma, \forall s<t$. 
Thus, the example set is always $\gamma$-strongly separable. 
\end{proof}
